Parametricity Consequences of GADTs Representations

### 2021-02-17 20-gadts/GADT-story.tex
added Enrico's text

\subsection{Algebraic Data Types}

@@

\begin{example}\label{ex:g1}
Use our favorite GADT to establish non-equivalence of the two reps
\end{example}

@@

{\color{red} ENRICO: Just write a section that can go here about
  1. for our favorite GADT.}

>>>>

\subsection{Algebraic Data Types}\label{sec:adts}

@@

\begin{example}\label{ex:g1}
{\color{red} THIS EXAMPLE IS ENRICO'S TEXT}
Consider the following GADT, defined in a Haskell-like syntax:
\begin{equation}\label{eq:example-GADT}
\begin{array}{l}
\mathsf{data }\; G : \ast \to \ast\;\mathsf{where}\\
\hspace*{0.4in}c\;:\;G\;1
\end{array}
\end{equation}
This GADT has a unique constructor, providing a data of type $G\;1$.
While we might say this is the simplest GADT possible, it still
provides an interesting case-study, showing the differences between
the primitive data type representation and the Church encoding
representation.

We now represent $G$ via its Church encoding.  Since $G$ has a single
constructor, which provides a data of type $G\,1$ without requiring
any input, its Church encoding is
\[
G\,\alpha = \mathsf{\forall f.\, (\forall a.\,
  f\,a) \to f\,\onet}
\]
meaning that $G\,A = \mathsf{Equal}\,\onet\,A$.  Thus, $G$ tests its
argument for equality against the unit type, $\onet$.

Recall from the previous section {\color{blue} use reference} that the
semantic interpretation of $\mathsf{Equal}\,\onet\,A$ is given by a
characteristic function, in the sense that it is the singleton 1 when
$A$ is 1, and the empty set otherwise.  Thus, as a data type
represented by Church encoding, G will be the data type containing one
(and only one) piece of data when instantiated on $\onet$, and
uninhabited otherwise.  This perfectly mirrors the syntax of $G$,
which indeed provides just one constructor when the type is
instantiated on 1, and none otherwise.  Thus, given a set $A$, the set
interpretation $G_0(A)$ is 1 if $A = 1$, and $\emptyset$ otherwise.

We now want a primitive representation of $G$.  Recall that $G$ has a
unique constructor $c : G\,\onet$, which can be equivalently expressed
as $c: \onet \to G\,\onet$.  By using a $\Lan$ type, we can express
$c$ as a constructor $(\Lan_{\onet} \onet)\alpha \to G \alpha$.  Thus,
as a primitive data type, we represent $G$ as $\mu \phi. \lambda
\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$.  {\color{blue}
  (WHERE?) notice that the constructors of a GADT have to be functors,
  because the GADT is the fixpoint of a higher order functor.}

The semantic interpretation of $\mu \phi. \lambda
\alpha. (\Lan^{\emptyset}_{\onet} \onet)\alpha$ is given by the
fixpoint of the functor yielded by the semantic interpretation of
$\Lan^{\emptyset}_{\onet} \onet$, the body of the $\mu$ type.  Notice,
though, that the recursive variable $\phi$ does not appear in the body
of the $\mu$ type.  Thus, the semantic interpretation is actually just
the semantic interpretation of the body itself.  Therefore, the
semantic interpretation is $\mathsf{\Lan}_1 1$, where $1$ is the
constantly $1$-valued functor $\ast \to \set$ from the category with a
single object $\ast$ to $\set$.

It turns out that $\mathsf{\Lan}_1 1$ is, in fact, none other than the
identity functor on $\set$.  Indeed, for any set $A$,
$(\mathsf{\Lan}_1 1) A$ is the set of all functions $a : 1 \to A$
modulo the smallest equivalence relation $\sim$ such that, for any $f
: 1 \to 1$, we have $a \sim a \circ f$.  But this is isomorphic to
$A$, because the only function $1 \to 1$ is the identity.
{\color{blue} Reference Computing left Kan extensions - M.R.Bush
  M.Leeming R.F.C.Walters}

The Church encoding representation of $G$ and the primitive
representation of $G$ are thus non-equivalent: while the former only
holds one data element in the instance $G\,1$ and none in other
instances, the latter is inhabited in all instances with the exception
of $G\,\emptyset$.  This makes sense if one remembers that the left
Kan extension, reflected into the syntactic type $\Lan$, is a functor,
and thus it comes with a functorial map.  This means that, for every
element $a \in A$, we can map the associated function $\bar{a} : 1 \to
A$ along $G$ to get a function $\map_{G}\,\bar{a} : G\,1 \to G\,A$,
and thus we have an element in $G\,A$ given by $\map_{G}\,\bar{a}\, c$
where $c \in G\,1$ is the element provided by the constructor.
\end{example}

@@

{\color{red} EVERYTHING BEFORE THE NOTES BELOW IS ENRICO'S TEXT:} To
even consider the question of whether our semantics is parametric we
first need to define a relational semantics for our types, i.e., a
relational interpretation for every type $G : \ast \to \ast$ such
that, for every relation $R : \rel(A, B)$, we have $G_1(R) :
\rel(G_0(A), G_0(B))$.

We shall show that this is not generally possible for the primitive
representation of data types, by considering the previously defined
type $G$ in Equation~\ref{eq:example-GADT}.  Recall that the semantic
interpretation of $G$ is given by $\mathsf{\Lan}_1 1$ which, in fact,
is none other than the identity functor on $\set$.

Analogously to the set interpretation, the relational interpretation
for $G$ is given by $\mathsf{\Lan}_{\Eq_1} \Eq_1$.  Notice that the
interpretation still essentially relies on a left Kan extension, but
this time it is an extension of $\rel$-valued functors.  We would need
to prove that, for every relation $R : \rel(A, B)$, we have that
$(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ is a relation on $(\mathsf{\Lan}_1
1) A$ and $(\mathsf{\Lan}_1 1)B$, i.e., $A$ and $B$.  Instead, we will
provide a counterexample showing that this is not true.

Consider the relation $R = (1, \emptyset, \emptyset)$.  We would thus
expect $(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ to be a relation with domain
$1$.  Because the left Kan extension preserves
projections~\cite{rie16} we can compute the domain as
$(\mathsf{\Lan}_{\Eq_1} 1) R$ (notice that the left Kan extension of
$1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is a functor $\rel
\to \set$).

Following {\color{blue} Reference Computing left Kan extensions -
  M.R.Bush M.Leeming R.F.C.Walters} we know that
$(\mathsf{\Lan}_{\Eq_1} 1) R$ is the set of all morphisms of relations
$f : \Eq_1 \to R$ modulo an equivalence relation; but there can be no
morphism of relations $\Eq_1 \to R$ because its second component would
have to be a function $1 \to \emptyset$, and there can be no such
function.  Thus, $(\mathsf{\Lan}_{\Eq_1} 1) R = \emptyset$, which is
not the domain we would have expected.

The above example shows that the set interpretation and the relational
interpretation for the primitive representation of a data type are
inconsistent with each other, and thus it is not even possible to
consider the question of whether the semantics is parametric.

One could regard the above example as artificial, as it employs an
empty relation on a non-empty domain.  Though, it is possible to apply
the same argument, with a slightly more complex calculation, even to a
total relation such as $(1, 2, 1 \times 2)$.

\vspace*{0.2in}

%%% CE are non functorial but parametric
As we have seen in the previous section {\color{blue} ref}, GADTs
represented as Church encodings do not necessarily have a map or a
functorial semantics.  In particular, this means that we cannot
possibly derive naturality results for those types, event though their
semantics is parametric.  Indeed, even though naturality is sometimes
seen as a consequence of parametricity, that is only true in case we
have a functorial semantics and types supporting a map.  Still, GADTs
represented as Church encodings can still exhibit consequences of
paramentricity other than naturality, as they do admit parametric
semantics {\color{blue} cite Bob's paper?}.

%%% Example with G
{\color{blue} In this paragraph, $\mathsf{G}$ and $\mathsf{t}$
  represent syntax, so should be sans-serif, but in the notes, the
  paragraph about set/relational interpretation and
  parametricity/IEL/AT uses serif font.}  For example, consider the
Church encoding of the GADT $\mathsf{G}$ defined
in~\ref{eq:example-GADT}.  If $\mathsf{t}$ is a term of type
$\mathsf{G}$ and $R$ is a relation on the sets $A$ and $B$, then
$t_0(A) \in G_0(A)$ and $t_0(B) \in G_0(B)$, and by the abstraction
theorem we have that $t_0(A)$ and $t_0(B)$ are related in $G_1(R)$.
We also know that $G_1(R)$ is the empty relation whenever $R$ is not
$\mathit{Equal}\,1$.  We thus deduce that there can be no term of type
$\mathsf{G\,a}$ unless $\mathsf{a}$ is semantically equivalent to
$\mathsf{1}$.  This result is the syntactic counterpart of the
inhabitation result for $\mathsf{G}$ {\color{blue} are inhabitation
  results about semantics or syntax?}  showing that all of the
instances of $\mathsf{G}$ are uninhabited, with the exception of the
instance for $\mathsf{1}$.  Of course, we could not possibly have a
term of an uninhabited type.

%%% Primitive representation of GADT is non parametric, but functorial
On the contrary, as we have shown previously {\color{blue} ref}, a
GADT represented as a primitive data type does not necessarily admit a
parametric semantics, but, as a fixpoint of a higher-order functor, it
does admit a functorial semantics and, consequently, a functorial map.
In particular, every type that has a functorial map also benefits from
naturality properties.  In other words, there is a suitable
rearrange-transform property as explained in Section~\ref{sec:adts}.

%%% Example with G
For example, $\mathsf{G}$ is associated to a functorial map,
$\map_{\mathsf{G}}$, such that, for every function $\mathsf{f : a \to
  b}$, we get a function $\map_{\mathsf{G}}\,\mathsf{f} : \mathsf{G\,a
  \to G\,b}$.  Then we have a term $\map_{\mathsf{G}}\,\mathsf{f\,c} :
\mathsf{G\,a}$ for every function $\mathsf{f : \mathsf{1} \to a}$,
where $\mathsf{c : G\,\mathsf{1}}$ is the constructor provided by the
definition of $\mathsf{G}$.
